---
title: "calibration_doc_appendix"
author: "K. Davidson"
date: "Last update: `r Sys.Date()`" 
output: pdf_document: df_output
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# libraries 
library(tidyverse)
library(egg)
library(readxl)
library(ggpubr)    # for stat_cor() in ggplot theme
library(scales)    # for pretty_breaks()
library(finalfit)  # for ff_glimpse() missing values summaries
library(naniar)    # for miss_var_summary()
library(car)       # for vif()
#library(glmmTMB)
library(mgcv)      # for gam() with regression splines
#library(gam)      # do not run if mgcv is loaded. for gam with loess
#library(lme4) 
library(AICcmodavg)
library(bbmle)     # for AICctab()
library(MuMIn)     # for dredge(), model.sel() etc.


# read in data
setwd("~/ANALYSIS/data")
cal.raw <- read_excel("calibration_2020_KDupdate.xlsx", sheet="Calibration_v3")        #***copy-pasted to wd() Jan18, may require updating***
options(scipen = 9999)


# cleaning & calculating 
cal <- cal.raw %>% 
  mutate(across(c("bankfull":"water_vis"), ~ifelse(.x=="NA", NA, .x))) %>%
  mutate_at(c("water_level", "water_discharge", "OE", "ground_cuml_carc_above", "peak_live_below", "ground_cuml_carc_above", "ground_cuml_carc_below", "aerial_unsexed_above", "aerial_unsexed_below", "ground_carcs_total_above", "hpe_adj"), as.numeric) %>%
  mutate_at(c("system_stability", "size", "water_clarity", "substrate_shade", "canopy_cover", "large_woody_debris", "lpe_method", "bankfull", "brightness", "cloud_cover", "fish_vis", "water_vis"), as.factor) %>%
  mutate(lpe = ifelse(lpe_method=="Aerial" & !is.na(aerial_unsexed_above), peak_live_above+aerial_unsexed_above, 
    ifelse(lpe_method=="Aerial" & is.na(aerial_unsexed_above), peak_live_above,
      ifelse(lpe_method=="Ground" & is.na(ground_cuml_carc_above), peak_live_above,
        ifelse(lpe_method=="Ground" & !is.na(ground_cuml_carc_above) & hpe_method!="MR", peak_live_above+ground_cuml_carc_above, 
          ifelse(lpe_method=="Ground" & !is.na(ground_cuml_carc_above) & hpe_method=="MR", peak_live_above, NA)))))) %>%
  mutate(index = hpe/lpe) %>%
  mutate(index_propn = lpe/hpe) %>%
  mutate(sid = paste(gazetted_stream_name, year, sep="-")) %>%
  mutate(usid = paste(gazetted_stream_name, year, lpe_method, sep="-")) %>%
  mutate(n_row = 1:nrow(.)) %>%
  mutate(size_recode = case_when(size=="V. Small"~1, size=="Small"~2, size=="Medium"~3, size=="Large"~4, size=="X-Large"~5)) %>%
  mutate(water_clarity_recode = case_when(water_clarity=="Clear"~1, water_clarity=="Pt. Turbid / Tannic"~2, water_clarity=="Turbid"~3)) %>%
  mutate(substrate_shade_recode = case_when(substrate_shade=="Light"~1, substrate_shade=="Medium"~2, substrate_shade=="Dark"~3)) %>%
  mutate(canopy_cover_recode = case_when(canopy_cover=="Low"~1, canopy_cover=="Medium"~2, canopy_cover=="High"~3)) %>%
  mutate(lwd_recode = case_when(large_woody_debris=="Low"~1, large_woody_debris=="Medium"~2, large_woody_debris=="High"~3)) %>%
  mutate(bankfull_recode = case_when(bankfull=="0-25"~1, bankfull=="25-50"~2, bankfull=="50-75"~3, bankfull=="75-100"~4)) %>%
  mutate(brightness_recode = case_when(brightness=="Dark"~1, brightness=="Medium"~2, brightness=="Bright"~3, brightness=="Full"~4)) %>%
  mutate(cc_recode = case_when(cloud_cover=="0"~1, cloud_cover=="25"~2, cloud_cover=="50"~3, cloud_cover=="75"~4, cloud_cover=="100"~5)) %>%
  mutate(brightness_recode = case_when(brightness=="Dark"~1, brightness=="Medium"~2, brightness=="Bright"~3, brightness=="Full"~4)) %>%
  mutate(fish_vis_recode = case_when(fish_vis=="Low"~1, fish_vis=="Medium"~2, fish_vis=="High"~3)) %>%
  mutate(water_vis_recode = case_when(water_vis=="0.25-0.5"~1, water_vis=="0.5-1.0"~2, water_vis=="1.0-3.0"~3, water_vis=="3.0-bottom"~4)) %>%
  mutate_at(c("water_clarity_recode", "substrate_shade_recode", "lwd_recode", "fish_vis_recode"), as.integer) %>%
  select(sid, usid, n_row, watershed_group:water_vis_recode)
cal$size <- factor(cal$size, levels=c(NA, "V. Small", "Small", "Medium", "Large", "X-Large"), ordered=T)
```

**Estimates and environmental conditions**    
Examining relationships between estimates, index, and environmental conditions. Will not be extensive to avoid over-eager data mining, moreso to examine trends that people have previously taken as a given (e.g., Index~discharge).

```{r include=F, echo=F}
water_plot <- function(x_var, y_var, ann_x, ann_y, ann_lab, x_lab, y_lab, colour){
  ggplot(cal, aes(x=.data[[x_var]], y=.data[[y_var]])) +
    annotate("text", x=ann_x, y=ann_y, label=ann_lab) +
    geom_point(shape=21, fill=colour, colour="black", stroke=1.5, size=3.5, alpha=0.8) +
    scale_y_continuous(limits=c(0, ann_y)) +
    labs(x=x_lab, y=y_lab) +
    theme_bw()
}
```

```{r message=F, warning=F, echo=F}
ggarrange(water_plot("water_discharge", "hpe", 0, 800000, "A", "Discharge", "HPE", "red"), 
          water_plot("water_discharge", "lpe", 0, 350000, "B", "Discharge", "LPE", "orange"), 
          water_plot("water_discharge", "index", 0, 15, "C", "Discharge", "Index", "yellow"), 
          water_plot("water_level", "hpe", 0, 350000, "D", "Level", "HPE", "green"), 
          water_plot("water_level", "lpe", 0, 125000, "E", "Level", "LPE", "blue"), 
          water_plot("water_level", "index", 0, 17, "F", "Level", "Index", "purple"), ncol=3, nrow=2)
```

*Fig 2. a) HPE and b) Index versus discharge (cms) and level (m; c and d, respectively).*

While there does not appear to be a relationship between overall abundance or index and water quantity, with the exception of perhaps Index~Discharge (Fig. 2c), this is not necessarily the case for individual systems (e.g., Stellako; Fig. 3a). Note though that the relationship holds only for discharge and not level (Fig. 3b). The relationship between level and discharge may be exponential, but this is driven by a small number of observations across several systems (Fig. 4c), and it is important to note that Environment Canada derives flow from level, likely based on an "ideal" exponential curve. Therefore only one of these variables should be used in modelling. 

```{r, echo=F, warning=F, include=F}
stellako_plot <- function(x_var, ann_x, ann_y, ann_lab, x_title){
  ggplot(data=cal%>%filter(gazetted_stream_name=="Stellako River"), aes(x=.data[[x_var]], y=index, group=lpe_method)) +
    geom_smooth(aes(colour=lpe_method), method="lm", alpha=0.2) +
    geom_point(aes(fill=lpe_method), shape=21, colour="black", size=3.5, alpha=0.7) +
    #scale_x_continuous(limits=c(4,24)) +
    #scale_y_continuous(limits=c(0.5,4.5)) +
    annotate("text", x=ann_x, y=ann_y, label=ann_lab) +
    labs(x=x_title, y="Index", fill="Stellako", colour="Stellako") +
    theme_bw() +
    theme(legend.position=c(0.1,0.7),
      legend.background = element_rect(colour=alpha("black", 0.5), fill=alpha("white",0.5)),
      legend.spacing.y = unit(0.1, "mm"),
      legend.spacing.x = unit(0.1, "mm"),
      legend.margin=margin(t=0.1, r=0.1, b=0.1, l=0.1, unit="cm"),
      legend.key.size = unit(0.2, "cm"),
      legend.text = element_text(size=7),
      legend.title = element_text(size=8))
}

flow_level<-ggplot(cal%>%filter(!is.na(water_level), !is.na(water_discharge)), aes(x=water_level, y=water_discharge, fill=gazetted_stream_name)) +
  annotate("text", x=0, y=1010, label="C") +
  geom_point(shape=21, colour="black", size=3.5, alpha=0.6) +
  labs(x="Level (m)", y="Discharge (cms)", fill="System") +
  theme_bw() +
  theme(legend.margin=margin(t=0.1, r=0.1, b=0.1, l=0.1, unit="cm"),
    legend.spacing.x = unit(0.1, "mm"),
    legend.spacing.y = unit(0.1, "mm"),
    legend.background = element_rect(colour=alpha("black", 0.5), fill=alpha("white",0.5)),
    legend.position=c(0.35,0.6),
    legend.text = element_text(size=7),
    legend.title = element_text(size=8),
    legend.key.size = unit(0.2, "cm")) +
  guides(fill=guide_legend(ncol=2))
```

```{r, echo=F, warning=F, message=F}
ggarrange(stellako_plot("water_discharge", 5, 4.1, "A", "Discharge (cms)"), stellako_plot("water_level", 0.1, 4.1, "B", "Level (m)"), flow_level, nrow=2, ncol=2)
```

*Fig 3. Relationship between a) river discharge (cms), b) river level (m) and the escapement index (HPE divded by LPE) for Stellako (1989-2019). c) Relationsip between discharge and level for paired locations, but note EC derives flow from level.*

<br>

<span style="text-decoration:underline"> A note on environmentals (APPENDIX) </span>    

```{r, echo=F, warning=F, message=F, include=F}
# Plot function for Index ~ Survey/stream type variables
env_plot_fx <- function(x_var, ann_x, ann_y, ann_lab, x_lab){
  ggplot(cal%>%filter(!is.na(.data[[x_var]]), !grepl("CANT", .data[[x_var]])), aes(x=.data[[x_var]], y=index)) +
    geom_point(shape=21, colour="black", size=3.5, fill="gray70", alpha=0.6) +
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se=F, colour="red", size=1, alpha=0.5) +
    geom_smooth(method = "loess", se=F, colour="blue", size=1, alpha=0.5) +
    annotate("text", x=ann_x, y=ann_y, label=ann_lab, fontface=2) +
    labs(x=x_lab, y="Index") +
    theme_bw() + 
    theme(legend.position=c(0.1,0.85),
      legend.background = element_rect(colour="black"),
      legend.spacing.y = unit(0.1, "mm"),
      legend.spacing.x = unit(0.1, "mm"),
      legend.margin=margin(t=0.1, r=0.1, b=0.1, l=0.1, unit="cm"),
      legend.key.size = unit(0.5, "cm"),
      axis.text=element_text(colour="black"))
  }
```

```{r, echo=F, warning=F, message=F}
ggarrange(env_plot_fx("bankfull", "0-25", 17, "A", "Bankfull (%)"), 
          env_plot_fx("water_vis", "0.25-0.5", 17, "B", "Water visibility (m)"), 
          env_plot_fx("size", "V. Small", 17, "C", "Stream size"), 
          env_plot_fx("OE", 25, 17, "D", "Observer efficiency (%)"), nrow=2, ncol=2)
```

*Fig 5. Relationships between the escapement index and a) percent bankfull, b) water depth visibility, c) stream size, and d) observer efficiency.* 








### 1. Environmental covariate competition ###

Using a sub-set of data, we can test the importance of core environmental covariates on each other to determine if static covariates are appropriate for full model testing. This subset was chosen based on observations where all data were recorded (Figure 6). This left a small dataset of n=39 calibration observations. Note that the core list of covariates could only be whittled down to n=6 covariates, so model over-fitting is likely.

```{r, echo=F, include=F}
##################
# COMPETE MODELS #
##################

#mg2 <- glm(index ~ 1 + bankfull + brightness + cloud_cover + fish_vis + OE, data=cal.env.train.new, family=Gamma(link = "inverse"), na.action=na.fail)
#mg2.dredge <- dredge(mg2, beta="none", rank="AICc", m.lim = c(1,4))
#dredge2.4 <- subset(mg2.dredge, delta < 4)



sc.cal.80 <- transform(cal.80, lpe=scale(lpe,center=FALSE))
m.global2 <- update(m.global, data=sc.cal.80)

options(scipen = 0)
summary(m.global2)

ss <- c(1, lpesc <- 1/attr(sc.cal.80$lpe,"scaled:scale"))

s_vals <- cbind(glm_unsc=suppressWarnings(sqrt(diag(vcov(m.global)$cond))),
                glm_sc=sqrt(diag(vcov(m.global2)$cond))*ss)
print(s_vals,digits=3)




#AICc(m.0, m.s, m.wc, m.b, m.cc, m.wv, m.oe)
#cand.models <- list(m.0, m.s, m.wc, m.b, m.cc, m.wv, m.oe)
#cand.names <- c("null", "size", "water clarity", "bankfull", "cloud cover", "water vis", "OE")
#t <- AICctab(m.0, m.s, m.wc, m.b, m.cc, m.wv, m.oe, nobs=39, logLik=T, base=T, weights=T, delta=T, sort=T)

```
